---
title: "OPEN & REPRODUCIBLE MICROBIOME DATA ANALYSIS SPRING SCHOOL 2018"
author: "Sudarshan"
date: "`r Sys.Date()`"
output: bookdown::gitbook
site: bookdown::bookdown_site
---

# Alpha diversities  

Alpha diversity measures are used to identify within individual taxa richness and evenness. The commonly used metrics/indices are Shannon, Inverse Simpson, Simpson, Gini, Observed and Chao1. These indices do not take into account the phylogeny of the taxa identified in sequencing. Phylogenetic diversity (Faith's PD) uses phylogenetic distance to calculate the diversity of a given sample.   

*It is important to note that, alpha diversity indices are sensitive to noise that is inherent to application of polymerase chain reaction and the  sequencing errors.*  

One has to consider the sequencing depth (how much of the taxa have been sampled) for each sample. If there is a large difference, then it is important to normalize the samples to equal sampling depth. First, we look at the sampling depth (no. of reads per sample).     

**Load packages**  
 

```{r, warning=FALSE, message=FALSE}

library(microbiome) # data analysis and visualisation
library(phyloseq) # also the basis of data object. Data analysis and visualisation
library(microbiomeutilities) # some utility tools 
library(RColorBrewer) # nice color options
library(ggpubr) # publication quality figures, based on ggplot2
library(DT) # interactive tables in html and markdown
library(data.table) # alternative to data.frame
library(dplyr) # data handling  

```

The data for tutorial is stored as *.rds file in the R project `phyobjects` folder.  

We will use the filtered phyloseq object from **Set-up and Pre-processing** section.  

```{r}

ps1 <- readRDS("./phyobjects/ps.ng.tax.rds")

# use print option to see the data saved as phyloseq object.

print(ps1)

```


```{r}

summary(sample_sums(ps1))

```

As is evident there is a large difference in the number of reads. Minimum is 1286 and maximum is 41694!! There is a ~30X difference!  

We can plot the rarefaction curve for the observed ASVs in the entire data set.  

```{r}

out_tab <- t(abundances(ps1))
p <- vegan::rarecurve(out_tab, 
                      step = 50, label = FALSE, 
                      sample = min(rowSums(out_tab), 
                                   col = "blue", cex = 0.6))

```

Not all samples are reaching a plateau and that few samples have high number of reads and high number of ASVs.  
Since we are comparing different body sites, some are expected to have low bacterial load.  
We will normalize to the lowest depth of at least 2000 reads to keep maximum samples for our anlaysis. This can be varied to remove samples with lower sequencing depth. This decision will depend on the research question being addressed.  

## Equal sample sums  

```{r}

set.seed(9242)  # This will help in reproducing the filtering and nomalisation. 

ps0.rar <- rarefy_even_depth(ps1, sample.size = 2000)

saveRDS(ps0.rar, "./phyobjects/ps0.rar.rds")

```

Check how much data you have now  

```{r}

ps0.rar <- readRDS("./phyobjects/ps0.rar.rds")


print(ps0.rar)

```


```{r}

 # quick check for sampling depth

barplot(sample_sums(ps0.rar), las =2)

```


```{r}
# quick check taxa prevalence

p.rar <- plot_taxa_prevalence(ps0.rar, "Phylum")

p.rar

```

Compare this to taxa prevalence plot from previous section of the tutorial.  

Do you see any difference?  

## Diversities  

### Non-phylogenetic diversities  

For more diversity indices please refer to [Microbiome Package](http://microbiome.github.io/microbiome/Diversity.html)  

Let us calculate diversity.  

```{r}

hmp.div <- diversities(ps0.rar, index = "all")

datatable(hmp.div)

```

This is one way to plot the data. 

```{r}
# get the metadata out as seprate object
hmp.meta <- meta(ps0.rar)

# Add the rownames as a new colum for easy integration later.
hmp.meta$sam_name <- rownames(hmp.meta)

# Add the rownames to diversity table
hmp.div$sam_name <- rownames(hmp.div)

# merge these two data frames into one
div.df <- merge(hmp.div,hmp.meta, by = "sam_name")

# check the tables
colnames(div.df)

# Now use this data frame to plot 
p <- ggboxplot(div.df, x = "scientific_name", y = "shannon",
              fill = "scientific_name", palette = "jco")
p + rotate_x_text()

colnames(hmp.div)
```


Alternative way  

```{r}

# convert phyloseq object into a long data format.  

div.df2 <- div.df[,c("scientific_name", "inverse_simpson", "gini_simpson", "shannon", "fisher", "coverage")]

# the names are not pretty. we can replace them 

colnames(div.df2) <- c("Location", "Inverse Simpson", "Gini-Simpson", "Shannon", "Fisher", "Coverage")

# check
colnames(div.df2)

div_df_melt <- reshape2::melt(div.df2)

head(div_df_melt)

```

The diversity indices are stored under column named *variable*.  

```{r, fig.height=8, fig.width=12}

# Now use this data frame to plot 
p <- ggboxplot(div_df_melt, x = "Location", y = "value",
              fill = "Location", 
              palette = "jco", 
              legend= "right",
              facet.by = "variable", 
              scales = "free")

p <- p + rotate_x_text() 
# we will remove the x axis lables

p <- p + rremove("x.text")
p
ggsave("./figures/Diversities.pdf", height = 4, width = 10)

```


```{r}

lev <- levels(div_df_melt$Location) # get the variables

# make a pairwise list that we want to compare.
L.pairs <- combn(seq_along(lev), 2, simplify = FALSE, FUN = function(i)lev[i])

pval <- list(cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1), symbols = c("****", "***", "**", "*", "n.s")) 

p2 <- p + stat_compare_means(comparisons = L.pairs, 
                             label = "p.signif", 
                             symnum.args = list(cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1), 
                                                symbols = c("****", "***", "**", "*", "n.s")))

print(p2)
```

### Phylogenetic diversity   

Phylogenetic diversity is calculated using the `picante` package.  

```{r, message=FALSE, warning=FALSE}

library(picante)

ps0.rar.asvtab <- as.data.frame(ps0.rar@otu_table)

ps0.rar.tree <- ps0.rar@phy_tree

# hmp.meta from previous code chunks

# We first need to check if the tree is rooted or not 

ps0.rar@phy_tree
# it is a rooted tree
df.pd <- pd(t(ps0.rar.asvtab), ps0.rar.tree,include.root=T) # t(ou_table) transposes the table for use in picante and the tre file comes from the first code chunck we used to read tree file (see making a phyloseq object section).


datatable(df.pd)

```

now we need to plot PD. Check above how to get the metadata file from a phyloseq object.  

```{r}
# now we need to plot PD

# We will add the results of PD to this file and then plot.

hmp.meta$Phylogenetic_Diversity <- df.pd$PD

```

**Plot**   

```{r}

pd.plot <- ggboxplot(hmp.meta, x = "scientific_name", 
                     y = "Phylogenetic_Diversity",
                     fill = "scientific_name", 
                     palette = "jco",
                     ylab = "Phylogenetic Diversity", 
                     xlab = "Body site",
                     legend = "right")
pd.plot <- pd.plot + rotate_x_text()

pd.plot + stat_compare_means(comparisons = L.pairs, 
                             label = "p.signif", 
                             symnum.args = list(cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1), 
                                                symbols = c("****", "***", "**", "*", "n.s")))

```

**NOTE:**  

There are arguments both for and against the use of rarefying to equal library size.  
The application of normalization method will depend on the type of research question. It is always good to check if there is a correlation between increasing library sizes and richness. Observed ASVs and Phylogenetic diversity can be affected by library sizes. It is always good to check for this before making a choice.    

```{r}

lib.div <- diversities(ps1, index = "all")

lib.div2 <- richness(ps1)


# let us add library size
lib.div$LibrarySize <- sample_sums(ps1)

lib.div$Richness <- lib.div2$`0`

colnames(lib.div)

ggscatter(lib.div, "LibrarySize", 
          "shannon") + 
  stat_cor(method = "pearson")

ggscatter(lib.div, "inverse_simpson", "LibrarySize",
          add = "loess") + 
  stat_cor(method = "pearson")

ggscatter(lib.div, "Richness", "LibrarySize",
          add = "loess") + 
  stat_cor(method = "pearson", 
           label.x = 100, 
           label.y = 50000)

```


```{r}

sessionInfo()

```


