---
title: "OPEN & REPRODUCIBLE MICROBIOME DATA ANALYSIS SPRING SCHOOL 2018"
author: "Sudarshan"
date: "`r Sys.Date()`"
output: bookdown::gitbook
site: bookdown::bookdown_site
---


# Composition plots  

Barplots are an easy and intuitive way of visualising the composition of your samples. However, the way this is implented in phyloseq causes problems with the order of the taxa in the legend at higher taxonomic levels.

We will use the filtered phyloseq object from the Microbiome Analytics Workshop 2018 tutorial I.  

**Load packages** 

```{r, warning=FALSE, message=FALSE}

library(microbiome) # data analysis and visualisation
library(phyloseq) # also the basis of data object. Data analysis and visualisation
library(RColorBrewer) # nice color options
library(ggpubr) # publication quality figures, based on ggplot2
library(dplyr) # data handling  

```

  
```{r}

ps1 <- readRDS("./phyobjects/ps1.rds")

# use print option to see the data saved as phyloseq object.

print(ps1)

```

## Barplot counts 

```{r, fig.height= 6, fig.width=20, warning=FALSE, message= FALSE}

ps1.com <- ps1

# We need to set Palette
taxic <- as.data.frame(ps1.com@tax_table)  # this will help in setting large color options

#colourCount = length(unique(taxic$Family))  #define number of variable colors based on number of Family (change the level accordingly to phylum/class/order)
#getPalette = colorRampPalette(brewer.pal(12, "Paired"))  # change the palette as well as the number of colors will change according to palette.

taxic$OTU <- rownames(taxic)  # Add the OTU ids from OTU table into the taxa table at the end.
colnames(taxic)  # You can see that we now have extra taxonomy levels.

taxmat <- as.matrix(taxic)  # convert it into a matrix.
new.tax <- tax_table(taxmat)  # convert into phyloseq compatible file.
tax_table(ps1.com) <- new.tax  # incroporate into phyloseq Object



# now edit the unclassified taxa
# tax_table(ps1.com)[tax_table(ps1.com)[, "Family"] == "f__", "Family"] <- "Unclassified family"


# We will also remove the 'f__' patterns for cleaner labels
# tax_table(ps1.com)[, colnames(tax_table(ps1.com))] <- gsub(tax_table(ps1.com)[, 
#    colnames(tax_table(ps1.com))], pattern = "[a-z]__", replacement = "")

# it would be nice to have the Taxonomic names in italics.
# for that we set this

guide_italics <- guides(fill = guide_legend(label.theme = element_text(size = 15, 
    face = "italic", colour = "Black", angle = 0)))


## Now we need to plot at family level, We can do it as follows:

# first remove the phy_tree

ps1.com@phy_tree <- NULL

# Second merge at family level 

ps1.com.fam <- aggregate_taxa(ps1.com, "Family", top = 10)

plot.composition.COuntAbun <- plot_composition(ps1.com.fam) + theme(legend.position = "bottom") + 
    scale_fill_brewer("Family", palette = "Paired") + theme_bw() + 
    theme(axis.text.x = element_text(angle = 90)) + 
  ggtitle("Relative abundance") + guide_italics + theme(legend.title = element_text(size=18))
  
plot.composition.COuntAbun

#ggsave("./Test_Outputfiles/Family_barplot_CountAbundance.pdf", height = 6, width = 8)


```


This plot is based on the reads/sample. You can see the reads are not evenly distributed over the samples, nevertheless their overall composition seems to be the same. The only thing that is different is the scaling. You don't need any other normalisation alogorithms. To check this in the next step we plot the relative abundance.

## Barplot relative abundance 

Make it relative abundance

```{r, fig.height= 6, fig.width=20, warning=FALSE, message= FALSE}

# the previous pseq object ps1.com.fam is only counts.

# Use traqnsform function of microbiome to convert it to rel abun.

ps1.com.rel <- microbiome::transform(ps1.com, "compositional")

ps1.com.fam2 <- aggregate_taxa(ps1.com.rel, "Family", top = 10)


plot.composition.relAbun <- plot_composition(ps1.com.fam2, 
                                             sample.sort = "scientific_name", 
                                             x.label = "env_material") + theme(legend.position = "bottom") + scale_fill_brewer("Family", palette = "Paired") + theme_bw() + 
    theme(axis.text.x = element_text(angle = 90)) + 
  ggtitle("Relative abundance") + guide_italics + theme(legend.title = element_text(size=18))
  
plot.composition.relAbun


#ggsave("./figures/Family_barplot_RelAbundance.pdf", height = 6, width = 8)


```

### Barplot customize 

```{r}

data.com <- plot.composition.relAbun$data
colnames(data.com)

```

```{r}


p.com <- ggplot(data.com, aes(x = Sample, y = Abundance, fill = OTU))
p.com <- p.com + geom_bar(position = "stack", stat = "identity")
p.com <- p.com + scale_x_discrete(labels = data.com$xlabel, breaks = data.com$Sample)
p.com <- p.com + facet_grid(~xlabel, scales = "free") + theme_bw()
p.com <- p.com + scale_fill_brewer("Family", palette = "Paired") 
p.com <- p.com + rremove("x.text") 

ggsave("./figures/Composition plots.pdf", height = 4, width = 6)


```

for more information [Microbiome tutorial](http://microbiome.github.io/microbiome/Composition.html)   


## Heatmaps

```{r}

# base plot
p.heat <- ggplot(data.com, aes(x = Sample, y = OTU)) + geom_tile(aes(fill = Abundance)) 

# Change color
p.heat <- p.heat + scale_fill_distiller("Abundance", palette = "RdYlBu") + theme_bw() 

# Make bacterial names italics
p.heat <- p.heat + theme(axis.text.y = element_text(colour = 'black', 
                                                    size = 10, 
                                                    face = 'italic')) 
# Make seperate samples based on main varaible
p.heat <- p.heat + facet_grid(~xlabel, 
                              scales = "free") + rremove("x.text") 

p.heat <- p.heat + ylab("Family")

#Clean the x-axis
p.heat <- p.heat + theme(axis.title.x=element_blank(),
                     axis.text.x=element_blank(),
                     axis.ticks.x=element_blank()) 

# Clean the facet label box
p.heat <- p.heat + theme(legend.key = element_blank(), 
                     strip.background = element_rect(colour="black", fill="white"))

print(p.heat)

ggsave("./figures/Heatmap.pdf", height = 4, width = 6)


# + geom_text(aes(label = round(Abundance)), size = 0.4)
```


**Extra**  

Following is an example of customizing the plot using ggpubr.  

```{r}


ps_df <- microbiomeutilities::phy_to_ldf(ps1.com, transform.counts = "compositional")

colnames(ps_df)
# this data.frame can be used to customize several plots.  

# example boxplot at phylum level

p.box <- ggstripchart(ps_df, "scientific_name", "Abundance", 
                      facet.by = "Phylum", color = "scientific_name",
                      palette = "jco")

p.box + rremove("x.text")

```




```{r}

sessionInfo()

```


